%!TEX root = ../template.tex
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% chapter4.tex
%% NOVA thesis document file
%%
%% Chapter with lots of dummy text
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\typeout{NT FILE chapter4.tex}%

\chapter{Conclusions}\label{cha:conclusions}

In this final chapter, some conclusions and future work will be presented.
This preparatory report focuses mostly on the theorectical part of the work, where first, \glspl{MPA} and their characteristics need to be understood, and then possible techniques to answer the proposed question are presented.
In~\ref{sec:fundamentals_of_mpa}, it was shown the advantages and limitations of \glspl{MPA}, as well as, their main performance parameters and the analysis of a basic rectangular \gls{MPA} was done.
Then, in~\ref{cha:stateoftheart_implementation}, multiple geometries and techniques for sizing, bandwidth enhancement, feeding and methods to acheive dual and tri-band operation were introduced, as well as, some use cases of multiband \glspl{MPA}.
Finally in that same section in~\ref{ai_optimization}, some \gls{AI} optimization techniques for the design of \glspl{MPA} where shown.
In~\ref{cha:workplan}, a timeline and work plan was presented, with tasks, which include the objectives defined in~\ref{sec:objectives}, to find the solution to the problem.

The future work is the major step to answer the proposed question. 
Starting with the definition of the specifications for the \gls{MPA} and its preliminary design. 
After this, the chosen design will be simulated to find the antenna's performance metrics and then optimized.
Once the final design is acheived, the fabrication of the antenna will proceed and real world measurements will be made. 
These measurements will then be compared with those obtained in the simulation to validate the design.